---
title: "Determinants of Housing Quality: Uncovering the Substantial Impact of Dwelling Type, Zoning Classification, and Lot Area on Overall Home Quality in Ames, Iowa"
author: "David Zeng"
format: html
editor: visual
---

## Abstract

The research question I planned to tackle was "What role do certain housing characteristics, specifically the type of dwelling, lot area, and zone classification, play in influencing the overall quality rating of a property?" The overall quality of homes significantly influences both market value and buyer satisfaction, making it a crucial factor in real estate analysis and housing market valuation. This analysis investigates the impact of three key variables—Dwelling Type, Zoning Classification, and Lot Area—on the overall quality of homes in Ames, Iowa. The hypothesis was that each of these factors would significantly affect overall home quality. Using a housing dataset from Kaggle with over 2,900 combined observations from Ames, Iowa, I selected the required variables to answer this question: MSSubClass, MSZoning, and LotArea, OverallQual, cleaned them into tidy format, and renamed them to make easier to understand: DwellType, ZoneClass, LotArea, and OverallQual. 3 separate ANOVA tests were performed to assess the significance of each variable's effect on overall home quality based on each category within their respective variables. The findings revealed that all three variables were statistically significant in affecting the overall quality of homes. More specific conclusions can be drawn from each individual variable's data analysis plot. These findings may prove useful for prospective home buyers, real estate professionals, and investors by providing insights that can lead to more informed decision-making in property valuation and investment. Understanding these factors allows for more accurate assessments of home quality, ultimately benefiting buyers, sellers, and investors alike.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r setup}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

message: false gets rid of any messages when loading a library during a render in a tab

don't print out the entire dataframe/dataset unless it is less than 15 rows
